### 2018-01-09 thesis/contents/reviews/multiplesclerosis.tex
changes with moji

\section{Multiple Sclerosis}

Multiple sclerosis (MS) is a chronic auto-immune demyelination disease of the central nervous system. It is a progressive disease characterized by wide-spread lesion, or plaques, within the brain, spinal cord, and optic nerve. Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability. MS symptoms can include cognitive deficits, fatigue, depression, upper and lower motor disabilities, and neuropathic pain. Currently there is no cure for MS. 
An estimated of up to 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}. The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.

\subsection{Classification}

MS is divided into four major groups: 1) Relapsing-remitting (RRMS), 2) Secondary progressive (SPMS), 3) Primary-progressive (PPMS), and 4) Progressive-relapsing (PRMS). RRMS is the most common form of MS, affecting about 85\% of the people with MS. It is characterized by sudden relapse (major flare-up) of symptoms, followed by periods of remission where symptoms disappear. SPMS may develop in some RRMS patients, where the symptoms worsen or level off. PPMS occur in 10\% of the MS population, where there are no remission periods, and symptoms will steadily worsen, with occasional plateaus. PRMS is the rarest form of MS, with less than 5\% of the MS population affected. It’s progressive in the beginning, with periods of relapse, it has no remission periods. 
Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) There is chronic inflammation of the CNS. One or more of these criteria result in a general diagnosis of MS. Both space and time dissemination criteria are confirmed with clinical MRI by radiologists. Detailed disability ratings are confirmed by the Expanded Disability Status Scale (EDSS)\cite{Kurtzke1983}, which tracks the progression of MS symptoms in an individual.

\subsection{Pathophysiology}

MS is defined by the existence of focal demyelination in the CNS that manifest in lesions or plaques. Ex-vivo dissections and ultra-structural examinations of this pathological tissue, obtained from autopsies of individuals who suffered from progressive MS or acute episodes. In acute tissue samples, the lesions showed signs of acute demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells. Partial remyelination by oligodendrocytes were also present, as well as signs of axonal swelling and injury. In chronic tissue samples, there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.

@@

The MS lesions or plaques are characterized as focal hyperintensities in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered when making a diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region requires at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

>>>>

\section{Multiple Sclerosis}
This chapter will review multiple sclerosis (MS) as a disorder, given the importance of MS-related TN as an independent TN subtype. MS is a chronic autoimmune demyelination disease of the central nervous system. It is a progressive disease characterized by wide-spread lesion, or plaques, within the brain, spinal cord, and optic nerve. Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability. MS symptoms can include cognitive deficits, fatigue, depression, motor weakness affecting the upper and lower extremities and neuropathic pain. Currently there is no cure for MS. An estimated 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}. The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.

MS is divided into four major clinical groups: 1) Relapsing-remitting (RRMS), 2) Secondary progressive (SPMS), 3) Primary-progressive (PPMS), and 4) Progressive-relapsing (PRMS). RRMS is the most common form of MS, affecting about 85\% of the people with MS. It is characterized by sudden relapse (major flare-up) of symptoms, followed by periods of remission where symptoms disappear. SPMS may develop in some RRMS patients, where the symptoms worsen or level off. PPMS occur in 10\% of the MS population, where there are no remission periods, and symptoms will steadily worsen, with occasional plateaus. PRMS is the rarest form of MS, with less than 5\% of the MS population affected. It is progressive in the beginning, with periods of relapse, it has no remission periods.

Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) Chronic inflammation of the CNS. One or more of these criteria result in a general diagnosis of MS. Both space and time dissemination criteria are confirmed with clinical MRI. Detailed disability ratings are confirmed by the Expanded Disability Status Scale (EDSS)\cite{Kurtzke1983}, which tracks the progression of MS symptoms in an individual.

MS is defined by the existence of focal demyelination in the CNS that manifests as lesions or plaques. Ultrastructural lesion analysis shows demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells. Partial remyelination by oligodendrocytes can be present, as well as signs of axonal swelling and injury. In tissue samples of longstanding MS there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.

@@

The MS lesions or plaques are characterized as focal hyperintensities in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered for the diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region requires at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

### 2018-01-09 thesis/contents/reviews/pain.tex
changes with moji

Trigeminal Neuralgia is a chronic facial pain syndrome. It is therefore imperative to review the evolving history and theories of pain. Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is generally described as having an unpleasant quality. Additionally, the experience of pain is complex, consisting of nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.

Humanity have long sought to understand pain. In antiquity, thinkers in classical Greek traditions, including Plato and Aristotle considered pain as primarily an emotion instead of a physical senses\cite{Moayedi2012a}. Hippocrates and Galen believed that pain arose from imbalances in vital fluids of the body, and considered the heart to be the organ responsible for the experience of pain \cite{Melzack1999}.

The earliest theory that attribute pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}). However, for Descartes, pain as an experience still could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul. For in my view exists only in the understanding. What I do explain is all the external movements which accompany this feeling in us." \cite[p. 148]{BookCottingham1991v3}. In his Treatise of Man, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch. In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time. When the entrance to the pore or small tube is opened in this way, the animal spirits from cavity enter and are carried through it...the fibres cause a movement in the brain which gives occasion for the soul...to have the sensation of pain" \cite[p.101-103]{BookCottingham1991v1}.

@@

Although the field of pain has changed significantly since the time of Descartes, a complete theory of pain still eludes modern neuroscience. Treatment of chronic pain is challenging still. However, investigations into unique syndromes such as trigeminal neuralgia offers an opportunity that further the understanding of pain and neuroscience at large.

@@

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes all the necessary information for pain. That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities is the  sensory pain experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\textit{Nociceptors}, the receptors in the peripheral nervous system that detect tissue damage can respond to a range of electrical or chemical signals released by noxious stimuli on the tissue, such as physical trauma or heat/cold temperatures exceeding the pain threshold. These receptors broadly separate into two categories, differentiated by the type of axonal innervation they receive. Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick or acute trauma. Dull pain is triggered by receptors that are innervated by slow conducting C fibres ($<1$ m/s). $A\delta$ and C fibres can be co-activated in the presence of noxious stimuli.

@@

Before reaching the neocortex, a majority of the second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopic organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.

@@

The intralaminar nuclei centromedian (CL) and parafascicular (Pf) receive noxious heat afferents. CL receive afferents from laminae V and VII, while Pf from laminae I and V. They project to the basal ganglia, motor and parietal cortices, and are thought to be involved in general attention, arousal, and motor functions.

@@

Pain percept is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}. The regions are not functionally independent, but rather thought to be dynamically inter-related.

Neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a} have revealed consistent regions of BOLD co-activations when undergone painful stimulation. These classic fMRI pain regions include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). 
Regions such as ACC, PFC and insula are also shown to be active in social pain through social exclusion, and emotional stress through PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).

@@

Other studies suggest pain is correlated to the dynamically changing patterns of activities over time in the regions involved. Whereby the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}, and the dynamic rhythms of activities that attune to and away from saliency forms a complex pain connectome \cite{Kucyi2015}. Conditions such as chronic pain therefore show a differentiated tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable nociceptive stimuli \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, which includes the location and intensity of pain include the primary sensory (S1) and secondary sensory (S2) cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. However, S1 has not been consistently reported in acute pain studies, possibily due to difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and is commonly found in pain studies. The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore it is believe to plays an important role in pain sensory-discrimination.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain procecssing, which includes the unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions . The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}.

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. DMN activation has been to correlate with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

There is a recent debate pertaining the dorsal posterior insula as the "pain cortex" \cite{Segerdahl2015a}. Segerdahl et al. suggested that there is strong coupling between absolute cerebral blood flow and tonic pain intensity, and utilized arterial spin-labelling to quantify cortical activation in response to pain stimulation. The contralateral dorsal posterior insula emerged as the only cortical region with significant correlation. Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is strong supportive findings of a homologue region in humans. Critics of this finding raised issues with the low number of subjects (n=7) and long sampling duration (45 s), and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.

@@

Chronic pain, as defined by the International Association for the Study of Pain (IASP), is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009}. It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months, other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}. There is a wide range of chronic pain conditions, including chronic back pain, migraine, arthritic pain, irritable bowel syndrome, phantom-limb pain, temporomandibular disorder, etc.

The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain. It is defined as "pain in one or more anatomical regions that persists or recurs for longer than 3 months, and is associated with significant emotional stress or significant functional disability" \cite{Treede2015}. It includes chronic pain conditions where the etiology is unknown. Conditions such as fibromyalgia, and irritable bowel syndrome fall under this category.

Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life.

@@

Neuropathic pain is a type of chronic pain, defined as "pain caused by a lesion or disease of the somatosensory nervous system" by the IASP. The pain experience may be constant or paroxysmal, related to a stimulus (evoked) or not-evoked. Animal models of neuropathic pain suggest that peripheral and spinal circuitry responsible for the transmission of nociception are altered under pathological conditions. The nervous system is believed to be altered by maladaptive plasticity following nervous system insult, and thereby lead to sustained neuropathic pain.

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission and processing circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation in the lower ganglions.

@@

 Studies have found that neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research.

 In general, brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain. These regions include the posterior insula, lateral thalamus and dorsal striatum. In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Lorenz2005}. In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}. Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and meidal prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}. A shift in forebrain activities is observed in neuropathic versus acute pain, abnormal activities in anterior cingulate cortex and prefrontal cortex is commonly observed in these studies \cite{Davis2006}.

>>>>

Trigeminal Neuralgia is a chronic neuropathic facial pain syndrome. The quest for the understanding of the pathophysiology of this syndrome requires a review of the history and theories of pain. Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is typically described as an unpleasant quality. The experience of pain is complex, consisting of interaction between nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.

Humanity have long sought to understand pain. In antiquity, thinkers in classical Greek traditions including Plato and Aristotle considered pain as primarily an emotion \cite{Moayedi2012a}. Hippocrates and Galen believed that pain arose from imbalances in vital fluids of the body, and considered the heart to be the organ responsible for the experience of pain \cite{Melzack1999}. These broad historical interpretations contrast more modern physiological approaches to the study of pain \cite{Melzack1999,Moayedi2012a}.

The earliest theory that attributes pain to a dedicated physical pathway is commonly credited to Ren\'{e} Descartes (\ref{fig:descarte}). However, for Descartes, pain as an experience  could not be divorced from the idea of the soul, as he had stated in his letter to the French polymath Marin Mersenne: "I do not explain the feeling of pain without reference to the soul. For in my view exists only in the understanding. What I do explain is all the external movements which accompany this feeling in us." \cite[p. 148]{BookCottingham1991v3}. In his \textit{Treatise of Man}, published in 1664, he stated: "Thus, for example, if fire A is close to foot B, the tiny parts of this fire...have the power also to move the area of skin which they touch. In this way they pull the tiny fibre which you see attach to it, and simultaneously open the entrance to the pore, located opposite the point where this fibre terminates -- just as when you pull one of a string, you cause a bell hanging at the other end to ring at the same time. When the entrance to the pore or small tube is opened in this way, the animal spirits from cavity enter and are carried through it...the fibres cause a movement in the brain which gives occasion for the soul...to have the sensation of pain" \cite[p.101-103]{BookCottingham1991v1}.

@@

@@

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain. That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities constitute the sensory component of the pain experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\textit{Nociceptors}, the receptors in the peripheral nervous system that detect tissue damage can respond to a range of electrical or chemical signals released by noxious stimuli on the tissue, such as physical trauma or heat/cold temperatures exceeding the pain threshold. These receptors broadly separate into two categories, differentiated by the type of axonal innervation they receive. Acute pain receptors are innervated by $A\delta$ fibres, which permits fast signal conductance (5 to 30 m/s) and trigger a prickly or sharp pain sensation; examples include needle prick other acute trauma. Dull pain is triggered by receptors that are innervated by slow conducting C fibres ($<1$ m/s). $A\delta$ and C fibres can be co-activated in the presence of noxious stimuli.

@@

Before reaching the neocortex, several important second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.

@@

@@

Pain percept, defined as the totality of the experience of pain, is thought to involve distributed regions of the brain, coined as the “pain neuromatrix”\cite{Melzack1999a}. The regions are not functionally independent, but rather thought to be dynamically inter-related.

 Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain were discovered through neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during painful stimulation.
Regions such as ACC, PFC and insula are also shown to be active in social pain during social exclusion and emotional stress associated with PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).

@@

A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved. Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}. The dynamic rhythms of activities that attune to and away from saliency, form a complex pain connectome \cite{Kucyi2015}. Conditions such as chronic pain therefore show a differentiated tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable nociceptive stimuli \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the primary sensory (S1) and secondary sensory (S2) cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies. The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore it is believe to plays an important role in pain sensory-discrimination.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain procecssing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}.

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. DMN activation has been correlated with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

There is a recent debate pertaining the dorsal posterior insula as the "pain cortex" \cite{Segerdahl2015a}. Segerdahl et al. suggested that there is strong coupling between absolute cerebral blood flow and tonic pain intensity, and utilized arterial spin-labelling to quantify cortical activation in response to pain stimulation. The contralateral dorsal posterior insula emerged as the only cortical region with significant correlation. Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is a strong supportive finding of a homologue region in humans. Critics of this finding raised issues with the low number of subjects, long sampling duration and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.

@@

Chronic pain is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009,Merskey2012}. It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months, other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}. The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain. It is defined as "pain in one or more anatomical regions that persists or recurs for longer than 3 months, and is associated with significant emotional stress or significant functional disability" \cite{Treede2015}. It includes chronic pain conditions where the etiology is unknown. Conditions such as fibromyalgia, and irritable bowel syndrome fall under this category. Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life \cite{May2008}.

@@

Neuropathic pain is a type of chronic pain, defined as "pain caused by a lesion or disease of the somatosensory nervous system"\cite{Merskey2012}. The pain experience may be constant or paroxysmal, related to a stimulus (evoked) or spontaneous. Animal models of neuropathic pain suggest that peripheral and spinal circuitry responsible for the transmission of nociception are altered under pathological conditions. The nervous system is believed to be altered by maladaptive plasticity following a nervous system insult thereby resulting in sustained neuropathic pain.

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission and processing circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation.

@@

 Neuropathic and acute nociceptive pain have a high degree of overlap in brain activations \cite{Moisset2007}, therefore differentiating between neuropathic and acute pain from neuroimaging continue to be an active area of research. Brain regions that are involved in the somatosensation of pain process are involved in both acute and chronic pain. These regions include the posterior insula, lateral thalamus and dorsal striatum. In chronic pain conditions however, there is increased activation in the anterior insula, ventral striatum, orbital frontal cortex and dorsal lateral prefrontal cortex \cite{Lorenz2005}. In chronic back pain, diffusion changes in fractional anisotropy was correlated with medial prefrontal cortex and nucleus accumbens connectivity in pain recovery, but not when the pain persistented \cite{Mansour2013}. Evidence also suggest that back pain chronification seem to increase activation in the emotional regions (amygada and medial prefrontal cortex) of the brain, and transition away from sensory region activations \cite{Hashmi2013}. A shift in forebrain activities is observed in neuropathic versus acute pain, abnormal activities in anterior cingulate cortex and prefrontal cortex is commonly observed in these studies \cite{Davis2006}.
